---
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Cygnus Example Analysis - A Quick Tutorial}
  %\usepackage[UTF-8]{inputenc}
---

## Cygnus Example Analysis - A Quick Tutorial

This tutorial covers reading a processed input file from AIVIA to generate an example report and 3D plots required for interactive data visualization. The instructions for package installation can be found <a href=https://yeinchung.github.io/Cygnus/>here</a>.

For this tutorial, we will use example data from the KP1.9 cell line derived extracellular vesicle imaging data. The data can be downloaded here.

```{r setup, include=FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(dpi=150, fig.width=4, fig.height=3) # Adjusted figure size
```

```{r echo=F, results='hide'}
library(Cygnus)
library(viridis)
library(plotly)
```

### Creating a Cygnus Object

To create a Cygnus object, specify metadata and markers columns from the input `.csv` file. You can use an interactive Shiny app or directly provide the column names.

```{r}
data.path <- "../inst/extdata/all_cells.csv" 
# cyg <- CreateCygnus(data.path) # Starts interactive app

cyg <- CreateCygnus(data.path,
                    markers_col = c("PanEV", "EpCAM", "MET",
                                    "SDC1", "EGFR", "ADAM10",
                                    "CTSH", "PDL1", "HER2"),
                    meta_col = "celltype")

cyg
```

### Adding Metadata

You can add metadata from another file using the following:

```{r}
cyg@ev_meta[['type']] <- "cell_line"
```

### Visualizing Marker Intensity Distribution

```{r}
plotDistribution(cyg)

plotDistribution(
    obj = cyg,
    plot_markers = "EpCAM",
    matrix = "Raw_Score",
    group_by = "celltype",
    group_colors = c("A549" = "#FF7966", "H2228" = "#CB8FF8", "KP" = "#7CC0F6")
)


```

### Scaling Expression Matrices

Differences in marker intensities may arise from antibody affinities rather than biology. Scaling helps normalize these differences.

```{r}
cyg <- scaleExpressionMatrix(cyg)
plotDistribution(cyg, matrix = "scaled_exp_matrix")
```

### Normalizing by PanEV

PanEV expression often correlates with EV size. This function normalizes marker intensities by PanEV expression.

```{r}
cyg <- normalizeByPanEV(cyg, "PanEV")
plotDistribution(cyg, matrix = "normalized_exp_mat") +
  theme(text = element_text(size = 8))

plotDistribution(
    obj = cyg,
    plot_markers = "EpCAM",
    matrix = "normalized_exp_mat",
    group_by = "celltype",
    group_colors = c("A549" = "#FF7966", "H2228" = "#CB8FF8", "KP" = "#7CC0F6")
    
)
```

### Creating a Binary Matrix

Binary matrices threshold marker expressions for classification purposes.

```{r}
cyg <- createBinaryMatrix(cyg, thresholds = 100)
plotDistribution(cyg, matrix = "binary_exp_matrix") +
  theme(text = element_text(size = 8))
```

### Visualizing Average Heatmaps

```{r}
plotAvgHeatmap(cyg, "celltype", scale = "row")
```

### Marker Analysis

Highlighting relevant markers for downstream analysis:

```{r}
cyg <- markRelevantMarkers(cyg, c("EpCAM", "MET",
                               "SDC1", "EGFR", "ADAM10",
                               "CTSH", "PDL1", "HER2"))
```

### Dimensionality Reduction and Visualization

#### PCA
```{r}
cyg <- runPCA(cyg, matrix_name = "normalized_exp_mat")
plotPCA(cyg, color_by = "celltype",plot_3d = T)
```

#### t-SNE
```{r}
set.seed(10)
cyg <- runTSNE(cyg)
plotTSNE(cyg, color_by = "celltype", marker_size = 6, plot_3d = T)
```

#### t-SNE with Normalized and Binary Matrices
```{r}
plotTSNE(cyg, color_by = "HER2", marker_size = 10, matrix_name = 'scaled_exp_matrix')
plotTSNE(cyg, color_by = "HER2", marker_size = 10, matrix_name = 'normalized_exp_mat')
plotTSNE(cyg, color_by = "HER2", marker_size = 10, matrix_name = 'binary_exp_matrix')
```

#### UMAP
```{r}
set.seed(100)
cyg <- runUMAP(cyg)
plotUMAP(cyg, color_by = "celltype", marker_size = 6, plot_3d = T)
```

### Clustering

Cluster EVs based on their expression profiles.

```{r}
cyg <- ClusterCygnus(cyg, n_clusters = 3)
#plotTSNE(cyg, color_by = 'k_means_clusters', marker_size = 10) 
```

### UpSet Plot for Marker Co-localization

Visualize marker intersections with an UpSet plot.

```{r}
# plotUpSet(cyg, markers = c("EpCAM", "MET",
#                            "SDC1", "EGFR", "ADAM10",
#                            "CTSH", "PDL1", "HER2"), 
#           nsets = 10, keep.order = TRUE)

colocal_marker <- getColocalizedMarkers(cyg)
plotUpset(cyg, colocal_marker)
```
